Cancer Applications of Choropleth Maps, and the Potential of Cartograms and

### 2019-12-18 review_article.Rmd
final proofing

---
title: "Cancer Applications of Choropleth Maps, and the Potential of Cartograms and Alternative Map Displays"
author: "Stephanie Kobakian, Jessie Roberts and Dianne Cook"
bibliography: reviewbib.bib
output:
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: no
    toc_depth: 3
    dev: jpeg
header-includes:
    - \usepackage{setspace}\doublespacing
    - \usepackage{fancyhdr}\pagestyle{fancy}\fancyhead[R]{\thepage}
link-citations: yes
fontsize: 11pt
geometry: margin = 1.2in
csl: springer-vancouver.csl
yolo: yes
---

@@

Cancer atlases communicate cancer statistics over geographic domains, typically with a choropleth map. They subdivide these domains into administrative regions such as countries, states, or suburbs. When communicating human-related statistics, the choropleth has a disadvantage in that it draws attention to sparsely populated rural areas to the neglect of small inner city areas. The smaller geographic areas are important to consider if they are densely populated. Alternative map displays, such as a cartogram or a hexagon tile map, can shift the attention of map users from the large rural areas by decreasing their size on the map display. This means alternative displays can be more effective at accurately communicating spatial patterns across spatial areas, by highlighting areas of interest. It is recommended that alternative displays are included in cancer atlases. In addition, with the ease of today's technology, user interaction with the displays is encouraged. Users should also be able to interactively display different statistics, such as incidence rate or relative incidence.

# Introduction
<!-- 436 words -->
<!-- What data are we working with, why, what are the purposes -->
Researchers, health authorities, governments, not-for-profits and the media are common communicators of cancer statistics. They often present statistics to the public as aggregated values for geopolitical areas. Presenting these statistics requires aggregating individual observations for the geographical units, especially for privacy protection, but also for political and policy purposes. Examples of typical geographical units include states, provinces, local government areas, and post/zip codes. It is easy to provide counts of the diagnoses or excess deaths within these areas. However, the amount of people living in each of the areas in a set of areas will be different. Calculating incidence rates involves dividing the count by the population for each area, for example, the rate per 100000 people. Counts and population datasets are collected routinely for public health reasons and may be made available to the general public as a service to the community.

<!-- Ways to visualize, whats our hypothesis -->
To visualize and communicate geospatial cancer statistics over geographic domain, a choropleth map is the common display. Choropleth maps show polygons representing the geographic units, where each polygon is shaded with a color according to the area-specific values of the statistic being conveyed. Visualizing this data is helpful as geographic patterns of disease may be obscured when reported in a table [@SAMGIS]. Providing a visual representation of cancer outcomes allows identification of geographic patterns of the disease that can then be addressed with public health policy and actions. The spatial distribution of the disease incidence can be examined using a choropleth and may reveal a trend in longitude or latitude, or rural vs urban, or coastal vs inland, or even specific hot spots of the disease. One of the key challenges with mapping spatial patterns of disease is the design of visualizations [@SE]. It is important to consider the strengths and weaknesses of designs, as visualizing diseases on maps is often the first step in exploratory spatial data analysis and helps in the formulation of hypotheses. This paper considers the current visualization techniques to communicate statistics to the public and their applications to cancer statistics. Alternative approaches are also considered, and the limitations of the visualization methods, highlighting the differences and historic use of these displays.

<!-- Structure of the paper -->
The paper is structured as follows. The next section describes the choropleth map, which is the common approach to disease maps and presents examples of atlases in use today. Section 3 discusses the limitations of the choropleth map. Section 4 describes alternative displays, including the cartogram, which is useful when the map has heterogeneously sized geographic units. Section 5 presents the limitations of the production and use of alternative displays. Disease maps are more useful when made interactive, Section 6 describes common options, along with a discussion of benefits and disadvantages.

# Traditional approaches for cancer map displays {#ch:choropleth_maps}
<!-- 381 words -->
<!-- what is a choropleth map -->
A choropleth map displays the geographical distribution of data over a set of spatial units by shading areas of a map [@EI, @BCM]. Faithful rendering of the geography, when combined with an appropriate color scheme, can reveal spatial patterns among data values. Identifying and explaining spatial structures, patterns, and processes involve considering the individuals and organizing them into representable units of communities [@SAMGIS]. Early versions of choropleth maps used symbols or patterns instead of color. Choropleth maps can be used for displaying disease data [@DMAHP], including cancer data [@CPISACA]. In epidemiology, choropleth maps are often used as a tool to study the spatial distribution of cancer incidence and mortality.

@@

## Cancer atlases {#ch:public}
<!-- 360 words -->
<!-- Displaying of cancer data on choropleth maps -->
A cancer atlas is a map, or collection of maps, representing cancer incidence and mortality for a country, or group of countries. Atlases are key to developing hypotheses regarding areas with unusually high rates, geographic correlations, work-related exposures, and high-risk diets [@MACM]. The data collection methods across regions and the administrative control within regions lends itself to choropleth visualization.Cancer maps and atlases dating back to Haviland's maps in 1875, and early work in US cancer atlases appearing in 1971 [@burbank]. The presentation of cancer statistics has increased with greater access to computational power and the availability of geographic information systems software [@SE].

<!-- Measures reported in cancer atlases -->
Cancer maps are effective tools for communicating incidence, survival, and mortality to a wide range of audiences, including the public and others not trained in statistical analysis. These visualizations enable non-expert audiences to interpret the outputs of sophisticated statistical analysis. Cruickshank (1947) as cited by S. D. Walter [@DMAHP], discusses using visuals as a 'formal statistical assessment of the spatial pattern'. Overwhelmingly, choropleth maps are visulisation chosen to communicate cancer statistics to members of the public and other non-expert audiences.

```{r choropleth-grid-create, message=FALSE, warning = FALSE, fig.cap = "A selection of choropleth cancer maps from online atlases that are publicly available. Maps of various countries were chosen: United Kingdom, Australia, Spain, USA, Canada, and display several different colour palettes and legends. These atllases are described in Table 1.", out.width='100%', fig.align='center', fig.width=12, fig.height=10}
library(cowplot)

@@

Table: 1<!--\label{tab:choropleths}--> A selection of choropleth cancer maps from online atlases.

@@

Epidemiologists and statisticians have developed the statistics used to communicate the burden of cancer over several decades. Table 2<!--\ref{tab:measures}--> summarizes the measures commonly presented in published cancer atlases. Mortality rates are commonly presented as relative rates of risk across the population and age-adjusted to correct for the higher prevalence of cancers in older populations. Stock developed the standardized mortality ratios through the 1930s, this is decribed by Howe [@HEDP].

Table: 2<!--\label{tab:measures}--> Common measures for reporting cancer information.

@@

<!-- 416 words -->
Roberts [@roberts2019communication] identified 33 cancer atlases published between 2010 and 2018. Each of these online atlases uses choropleth maps. All except one of these were published by non-commercial organizations, including not-for-profits, government, research organizations, advocacy groups or government-funded partnerships. Figure 1<!--\ref{fig:choropleth-grid-create}--> displays a subset of maps from these atlases, the selection varies in the geographies explored. Figure 1<!--\ref{fig:choropleth-grid-create}-->b shows Globocan 2018 [@Globocan] which explores Estimated Cancer Incidence, Mortality and Prevalence Worldwide using data sourced from cancer registries of each country. The Bowel Cancer Australia Atlas in Figure 1<!--\ref{fig:choropleth-grid-create}-->d presents an example of a cancer specific atlas, it shows the average Standardized Incidence Ratio of colorectal cancer for Australian males from 2006 to 2010 in Australia [@Bowel]. Like many of the atlases examined, there is a choice of gender displayed in the Bowel Cancer Atlas. An alternative approach used in the Map of Cancer Mortality Rates in Spain, Figure 1<!--\ref{fig:choropleth-grid-create}-->f shows side-by-side maps of the relative risk of lung cancer separately for men and women [@cancerSpain].

There are vast differences in the geographies considered by online atlases. 
Figure 1<!--\ref{fig:choropleth-grid-create}-->b shows global age-standardized incidence rates (per 100,000) for all invasive cancers for both men and women, aggregated at a national level for 2018. The United States Cancer Statistics [@USInteractive] presented in Figure 1<!--\ref{fig:choropleth-grid-create}-->e contains the incidence rate per 100,000, of all cancer types for men and women in the United States in 2016, aggregated at the state level. The Environment and Health Atlas of England and Wales, Figure 1<!--\ref{fig:choropleth-grid-create}-->a shows the relative risk for women developing lung cancer in England and Wales in 2010 [@EnvEnglandWales2] at a neighborhood (small-area) scale. The Atlas of Cancer in Queensland, Figure 1<!--\ref{fig:choropleth-grid-create}-->c shows the relative incidence ratio of lung cancer in males for each Statistical Area at Level 2 [@abs2016] in the state of Queensland within Australia [@QLDcancerAtlas].

Age-specific atlases are less common, Figure 1<!--\ref{fig:choropleth-grid-create}-->g displays Atlas of Childhood Cancer in Ontario, this communicates the incidence rate of childhood cancers per 100,000 (by census division) for children aged 0-14, in Ontario from 1995 to 2004 [@OntarioPediatric]. This atlas communicates information regarding a small geographic region.

## Additional considerations
<!-- 479 words -->
Cancer atlases often display supplementary graphs and plots to add more information. Additional materials such as tables, graphs, and text explanations support understanding and inference derived from maps, ensuring the message communicated will be consistent across a range of viewers [@CPISACA]. The many displays of statistical summaries, including dot plots, bar plots, box plots, cumulative distribution plots, scatter plots, and normal probability plots, can provide alternative views of the cancer statistics. These can also display supporting statistics such as error, confidence intervals, distributions, sample or population sizes, and standard deviation.

The statistics communicated in atlases are often used to describe differences between areas. This can occur at different levels of aggregation. Aggregation of global health statistics occurs within administrative and arbitrarily defined regions, such as those used by the World Health Organization and the United Nations [@IARC_3]. World atlases can allow for displays of data aggregated into continents, countries, states, provinces and congressional districts [@USInteractive]. Each population area will probably have a different number of people. Cancer atlases may also communicate the distribution of the population living in all areas in a table or histogram display [@NICR_1]. Atlases can connect the population to the land available to them by communicating population density. Section  4<!--\ref{ch:alternatives}--> considers alternatives to choropleth that introduce population into the visualizations.

Map users can define the population further using the demographics of the individuals within each area. Demographics that can influence cancer rates include information regarding the age and sex distribution of the areas presented. Some digital atlases surveyed allow users to control the display to select subsets such as males, females or those aged over 65. Similarly, socioeconomic indicators, such as unemployment rates, poverty rates, remoteness, and education levels, can explain how the experience of cancer prevalence varies for various members of society. Few atlases provide this level of detail.

Introducing population and demographic information helps to interpret the rates in areas effectively, but there will still be uncertainty around the rates. To address this, a cancer atlases often communicates uncertainty about the value of a statistic. There are several potential sources of uncertainty: sampling error, errors arising from the disease reporting process (or data collection), and errors arising from the statistical modeling or simulation process. The most common measures used to represent uncertainty were credible or confidence intervals (CIs). Displaying the uncertainty associated with reported statistics is a vital feature of a cancer map, but it is difficult to display effectively. The map focuses on displaying the statistic and lacks additional space to represent the uncertainty. Providing an adjacent map or overlaying maps with symbols [@VSSDCUC] are two approaches.

# Limitations of choropleth displays {#ch:chorolimit}

@@

# Contemporary alternatives to choropleth maps {#ch:alternatives}

## Cartograms
<!-- 1496 words inc table-->

@@

<!-- Denominators -->
When a map creator desires a uniform population density of the map base, the purposeful distortion of the map space is beneficial. The "population distribution is often extremely uneven in former British colonies" [@ACTUC], and this makes the distortion necessary, [@CTTMB] population will then become a uniformly distributed background for the statistic presented [@GOINO]. An area cartogram [@NAC], or population-by-area cartogram [@TAAM] occurs after the distortion of the geographical shape according to population. Event cartograms change the area of regions on a map depending on the amount of disease-related events, but this does not consider the effects of land area and population [@VSSDCUC].

<!-- Why transform?-->
<!-- Common variables used to create cartograms, e.g. population, mortality -->
Cartograms provide an alternative visualization method for statistical and geographical information. The key difference between a choropleth and a cartogram is the desirable augmentation of the size, shape, or distance of geographical areas [@ACTUC]. Monmonier [@HTLWM] suggests that map creators can use white lies to create useful displays by distorting the geometry and suppressing features. It is easy for the average person to disregard the impact of transformations used to create cartograms. One disadvantage of the conventional map is interpreting spatial patterns is difficult because of the sparsely populated rural areas being emphasized in a choropleth map, and the inner city areas are geographically small. The distortion of each area when making a cartogram is proportional to a value other than the actual earth size area [@NAC]. The distortion in an area cartogram accounts for the population density, preventing it from obscuring the spatial patterns [@TAAM]. The spatial transformation of map regions relative to the data emphasizes the data distribution instead of land size [@CBATCC]. When visualizing population statistics, Dorling [@ACTUC] considers this design 'more socially just', or honest [@NISCC], giving equitable representation and attention to all members of the population and reducing the visual impact of large areas with small populations [@DMAHP]. Howe [@HEDP] suggests that 'cancer occurs in people, not in geographical areas' and that spatial socio-economic data, like cancer rates, are best presented on a cartogram for urban areas as the population map base avoids allocating 'undue prominence' to rural areas [@CTTMB].

@@

```{r ggcartograms, fig.cap = "Common alternatives to maps, showing the same information for the United States of America: (a) contiguous cartogram, (b) non-contiguous, shape-preserved cartogram, (c)  Dorling cartogram (non-contiguous), (d) hexagon tilemap (non-contiguous). Maps (a) - (c) are created by resizing and reshaping the states of the USA to match the 2015 population of the state. This provides a better sense of the extent of disease relative to the population in the country and can help ease losing information about physically small but population-dense states. Map creators give each state equal size and thus equal emphasis in (d) the hexagon tile map.", message=FALSE, warning=FALSE, out.width='100%'}
g1 <- png::readPNG("figures/ggcont.png")
g2 <- png::readPNG("figures/ggncont.png")
g3 <- png::readPNG("figures/ggdorl.png")
g4 <- png::readPNG("figures/gghexmap.png")
#knitr::include_graphics("figures/usa_grid.png")

@@

Table: 3<!--\label{tab:usa}--> Maps used to present statistics for the United States of America. The colour of each state communicates the average age-adjusted rate of incidence for lung and bronchus for females and males in the United States 2012-2016.

@@

### Contiguous

@@

A contiguous cartogram alters the choropleth according to a statistic and maintains connectivity of the map regions. Min Ouyang and Revesz [@ACA] present three algorithms for creating value-by-area cartograms. They implement 'map deformation' to account for the value assigned to each area. Other methods include Tobler’s Pseudo-Cartogram Method, Dorling’s Cellular Automaton Method [@ACTUC], Radial Expansion Method of Selvin et al., Rubber Sheet Method of Dougenik et al., Gusein-Zade and Tikunov’s Line Integral Method, Constraint-Based Method (Kocmoud and House) [@CBATCC].

Figure 2<!--\ref{fig:ggcartograms}--> a shows a population contiguous cartogram of the United States. All states are visible and the shape of the United States overall is still recognizable. In contrast, Figure 3<!--\ref{fig:auscartograms}--> a shows an Australian contiguous cartogram also based on population. The south east is completely distorted, areas with low population are still large very the map, as their initial area was used to determine the area allocated in the contiguous cartogram.

@@

### Non-contiguous

Non-contiguous cartograms prioritize the shapes of the areas instead of connectivity. Each area stays in a similar position to its location on a choropleth map. Displaying the choropleth map base allows map users to make comparisons regarding the change in the area. The addition is the gap between areas, created as each area shrinks or grows according to the associated value of the statistic. Olson [@NAC] discusses the creation of these maps and the significance of the empty areas left between the geographic boundaries and the new shape. It is the 'difference from the original map that is the real message' of these displays.

The distortion of region shapes on the contiguous cartogram presents an additional hurdle to visual recognition, and the non-contiguous cartogram eliminates this hurdle, and presents the meaningful empty-space property [@ECGC, @NAC].

### Dorling

Daniel Dorling presents an alternative display engineered to highlight the spatial distribution and neighborhood relationships without complex distortions of borders and boundaries. Daniel Dorling puts forward a simple question and his approach opposes preserving the intricate shape details [@ACTUC]:

@@

He acknowledged the sophistication of contiguous cartograms but critiqued their 'very complex shapes,' he answers this with his implementation of maps created using 'the simplest of all shapes'. Circular cartograms use the same simple shapes for every region represented and resize the shapes according to the statistic represented or the population for a base map. This familiar shape may be more effective for understanding the spatial distribution than contiguous cartograms. Contiguous cartograms use a gravity model to apply distortions to create the 'nonsense' shapes that have 'no meaning' [@NISCC]. This method applies a gravity model to produce a compelling map, that avoids overlaps and keep spatial relationships with neighboring areas over many iterations. This implementation can work for up to ‘one hundred thousand’ areas.

Raisz laid the groundwork for this approach in the mid-1930s, and his rectangular cartograms provide dramatic comparisons and are especially useful for correcting misconceptions communicated by geographic maps. @TFYCC quotes the official definition of Value-Area Cartograms, the simplistic displays which represent each area as a single rectangle, sized according to the value of the statistic. This rectangular display may sacrifice contiguity but allows for tiling where geographic neighbors placed in suitable relative positions also share borders [@CDWCS]. Rectangular cartograms communicate bi-variate displays of the population by the size of each rectangular, and they use color to communicate a second variable [@ORC].

## Tile Map
<!-- 173 words -->
A tile map provides a tessellated display of consistent shapes.  A similar method to a rectangular cartogram, represents each geographic area using a square. The squares are tessellated to create a grid. Each area is represented by a square of the same dimensions, each tile is usually one unit of measurement, this could be geographic regions such as states or population-based that use a consistent measure of population for each tile. Regions with over four neighbors require some necessary displacement. The tile map uses color to represent a value of a statistic for each area. A similar method to a rectangular cartogram represents each geographic area using a square of the same dimensions. There are online media sources using this method, these include [NPR](https://www.npr.org/2016/10/18/498406765/npr-battleground-map-hillary-clinton-is-winning-and-its-not-close), [FiveThirtyEight](https://projects.fivethirtyeight.com/2016-election-forecast/), [The Wall Street Journal](http://graphics.wsj.com/elections/2016/2016-electoral-college-map-predictions/), [The Washington Post](https://www.washingtonpost.com/graphics/politics/2016-election/50-state-poll/). Tile maps may be difficult to create as they are best created manually, they require additional time and care as the number of geographic areas to include increases.
<!-- cite mosaic cartograms -->

>>>>

---
title: Cancer Applications of Choropleth Maps, and the Potential of Cartograms and
  Alternative Map Displays
author: Stephanie Kobakian, Jessie Roberts and Dianne Cook
csl: springer-vancouver.csl
output:
  word_document:
    toc: no
    toc_depth: '3'
  pdf_document:
    dev: jpeg
    fig_caption: yes
    number_sections: yes
    toc: no
    toc_depth: 3
geometry: margin = 1.2in
header-includes:
- \usepackage{setspace}\doublespacing
- \usepackage{fancyhdr}\pagestyle{fancy}\fancyhead[R]{\thepage}
link-citations: yes
fontsize: 11pt
bibliography: reviewbib.bib
yolo: yes
---

@@

Cancer atlases communicate cancer statistics over geographic domains, typically with a choropleth map. They subdivide these domains into administrative regions such as countries, states, or suburbs. When communicating human-related statistics, the choropleth has a disadvantage in that it draws attention to sparsely populated rural areas to the neglect of small inner city areas. The smaller geographic areas are important to consider if they are densely populated. Alternative map displays, such as a cartogram or a hexagon tile map, can shift the attention of map users from the large rural areas by decreasing their size on the map display. This means alternative displays can be more effective at accurately communicating spatial patterns across spatial areas. It is recommended that alternative displays are included in cancer atlases. In addition, with the ease of today's technology, user interaction with the displays is encouraged. Users should also be able to interactively display different statistics, such as incidence rate or relative incidence, or filtered by demographic variables.

# 1. Introduction
<!-- 436 words -->
<!-- What data are we working with, why, what are the purposes -->
Researchers, health authorities, governments, not-for-profits and the media are common communicators of cancer statistics. They often present statistics to the public as aggregated values for geopolitical areas. Presenting these statistics requires aggregating individual observations for the geographical units, especially for privacy protection, but also for political and policy purposes. Examples of typical geographical units include states, provinces, local government areas, and post/zip codes. It is easy to provide counts or incidence rates of the diagnoses of these areas.  This type of data is routinely collected for public health reasons and may be made available to the general public as a service to the community.

<!-- Ways to visualize, whats our hypothesis -->
To visualize and communicate geospatial cancer statistics over geographic domain, a choropleth map is the common display. Choropleth maps show polygons representing the geographic units, where each polygon is shaded with a color according to the area-specific values of the statistic being conveyed. Visualizing this data is helpful as geographic patterns of disease may be obscured when reported in a table [@SAMGIS]. Providing a visual representation of cancer outcomes allows identification of geographic patterns of the disease that can then be addressed with public health policy and actions. The spatial distribution of the disease incidence can be examined using a choropleth and may reveal a trend in longitude or latitude, or rural vs urban, or coastal vs inland, or even specific hot spots of the disease. One of the key challenges with mapping spatial patterns of disease is the design of visualizations [@SE]. It is important to consider the strengths and weaknesses of designs, as visualizing diseases on maps is often the first step in exploratory spatial data analysis and helps in the formulation of hypotheses. This paper considers the current visualization techniques to communicate statistics to the public and their applications to cancer statistics. Alternative approaches are posed because they may be more effective than contemporary techniques. The limitations of the visualization methods, highlighting the differences and historic use of these displays is discussed.

<!-- Structure of the paper -->
The paper is structured as follows. The next section describes the choropleth map, which is the common approach to disease maps and presents examples of atlases in use today. Section 3 discusses the limitations of the choropleth map. Section 4 describes alternative displays, including the cartogram, which is useful when the map has heterogeneously sized geographic units. Section 5 presents the limitations of the production and use of alternative displays. Disease maps are more useful when made interactive, and common options are described in Section 6, along with a discussion of benefits and disadvantages.

# 2. Traditional approaches for cancer map displays {#ch:choropleth_maps}
<!-- 381 words -->
<!-- what is a choropleth map -->
A choropleth map displays the geographical distribution of data over a set of spatial units by shading areas of a map [@EI, @BCM]. Faithful rendering of the geography, when combined with an appropriate color scheme, can reveal spatial patterns among data values. Identifying and explaining spatial structures, patterns, and processes involve considering the individuals and organizing them into representable units of communities [@SAMGIS]. Early versions of choropleth maps used symbols or patterns instead of color. Choropleth maps can be used for displaying disease data [@DMAHP], including cancer data [@CPISACA]. In epidemiology, choropleth maps are often used as a tool to study the spatial distribution of cancer incidence and mortality.

@@

## 2.1 Cancer atlases {#ch:public}
<!-- 360 words -->
<!-- Displaying of cancer data on choropleth maps -->
A cancer atlas is a map, or collection of maps, representing cancer incidence and mortality for a country, or group of countries. Atlases are key to developing hypotheses regarding areas with unusually high rates, and geographic correlations [@MACM]. The data collection methods across regions and the administrative control within regions lends itself to choropleth visualization. Cancer maps and atlases date back to Haviland's maps in 1875, and early work in US cancer atlases appearing in 1971 [@burbank]. The presentation of cancer statistics has increased with greater access to computational power and the availability of geographic information systems software [@SE].

<!-- Measures reported in cancer atlases -->
Cancer maps are effective tools for communicating incidence, survival, and mortality to a wide range of audiences, including the public and others not trained in statistical analysis. These visualizations enable non-expert audiences to interpret the outputs of sophisticated statistical analysis. Cruickshank (1947) as cited by S. D. Walter [@DMAHP], discusses using visuals as a 'formal statistical assessment of the spatial pattern'. Overwhelmingly, choropleth maps are visualisations chosen to communicate cancer statistics to members of the public and other non-expert audiences.

```{r choropleth-grid-create, message=FALSE, warning = FALSE, fig.cap = "Figure 1: A selection of choropleth cancer maps from online atlases that are publicly available. Maps of various countries were chosen: United Kingdom, Australia, Spain, USA, Canada, and display several different colour palettes and legends. These atllases are described in Table 1.", out.width='100%', fig.align='center', fig.width=12, fig.height=10}
library(cowplot)

@@

Table: Table 1: <!--\label{tab:choropleths}--> A selection of choropleth cancer maps from online atlases.

@@

Epidemiologists and statisticians have developed the statistics used to communicate the burden of cancer over several decades. Table 2<!--\ref{tab:measures}--> summarizes the measures commonly presented in published cancer atlases. Mortality rates are commonly presented as relative rates of risk across the population and age-adjusted to correct for the higher prevalence of cancers in older populations. As described in Howe [@HEDP], Englishman P. Stocks advanced the field of mortality statistics by introducing the standardized mortality ratios in the 1930s, which is an improvement on crude death rates.

Table: Table 2: <!--\label{tab:measures}--> Common measures for reporting cancer information.

@@

<!-- 416 words -->
Roberts [@roberts2019communication] identified 33 cancer atlases published between 2010 and 2018. Each of these online atlases uses choropleth maps. All except one of these were published by non-commercial organizations, including not-for-profits, government, research organizations, advocacy groups or government-funded partnerships. Figure 1<!--\ref{fig:choropleth-grid-create}--> displays a subset of maps from these atlases, the selection varies in the geographies explored. Figure 1<!--\ref{fig:choropleth-grid-create}-->b shows Globocan 2018 [@Globocan] which explores Estimated Cancer Incidence, Mortality and Prevalence Worldwide using data sourced from cancer registries of each country. The Bowel Cancer Australia Atlas in Figure 1<!--\ref{fig:choropleth-grid-create}-->d presents an example of a cancer specific atlas -- it shows the average Standardized Incidence Ratio of colorectal cancer for Australian males from 2006 to 2010 [@Bowel]. Like many of the atlases examined, there is a choice of gender displayed in the Bowel Cancer Atlas. Gender is displayed in side-by-side maps in the Map of Cancer Mortality Rates in Spain (Figure 1f)<!--\ref{fig:choropleth-grid-create}--> [@cancerSpain].

Resolution of the maps varies greatly. Figure 1<!--\ref{fig:choropleth-grid-create}-->b shows global information at a national level. The United States Cancer Statistics [@USInteractive] shows data aggregated at the state level. The Environment and Health Atlas of England and Wales [@EnvEnglandWales2] (Figure 1<!--\ref{fig:choropleth-grid-create}-->a) shows the relative risk for women developing lung cancer  at a neighborhood (small-area) scale. The Atlas of Cancer in Queensland (Figure 1<!--\ref{fig:choropleth-grid-create}-->c) shows the relative incidence ratio of lung cancer in males for each Statistical Area at Level 2 [@abs2016] in the state of Queensland within Australia [@QLDcancerAtlas].

Age-specific atlases are less common. Figure 1<!--\ref{fig:choropleth-grid-create}-->g displays Atlas of Childhood Cancer in Ontario, this communicates the incidence rate of childhood cancers per 100,000 (by census division) for children aged 0-14, in Ontario from 1995 to 2004 [@OntarioPediatric].

## 2.2 Additional considerations
<!-- 479 words -->
Cancer atlases often display supplementary graphs and plots to add more information. Additional materials such as tables, graphs, and text explanations support understanding and inference derived from maps, ensuring the message communicated will be consistent across a range of viewers [@CPISACA]. The many displays of statistical summaries, including dot plots, bar plots, box plots, cumulative distribution plots, scatter plots, and normal probability plots, can provide alternative views of the cancer statistics. These can also display supporting statistics such as error, confidence intervals, distributions, sample or population sizes, and standard deviation.

The statistics communicated in atlases are often used to describe differences between areas. This can occur at different levels of aggregation. Aggregation of global health statistics occurs within administrative and arbitrarily defined regions, such as those used by the World Health Organization and the United Nations [@IARC_3]. World atlases can allow for displays of data aggregated into continents, countries, states, provinces and congressional districts [@USInteractive]. Each population area will probably have a different number of people, which is typically used to calibrate the statistic. Cancer atlases may also communicate the distribution of the population living in all areas in a table or histogram display [@NICR_1]. Atlases can connect the population to the land available to them by communicating population density.

Maps can also be used to focus on demographic strata, such as age and sex. Some of the digital atlases surveyed allow subsets such as males, females, or those aged over 65, to be selected for display. Similarly, socioeconomic indicators, such as unemployment rates, poverty rates, remoteness, and education levels, can be used to filter data, in order to communicate how cancer prevalence varies for different members of society. Few atlases provide this level of detail.

Introducing population and demographic information helps to interpret the rates in areas effectively, but there will still be uncertainty around the rates. To address this, a cancer atlas often communicates uncertainty about the value of a statistic. There are several potential sources of uncertainty: sampling error, errors arising from the disease reporting process (or data collection), and errors arising from the statistical modeling or simulation process. The most common measures used to present uncertainty are credible or confidence intervals (CIs). Displaying the uncertainty associated with reported statistics is a vital feature of a cancer map, but it is difficult to display effectively. The map focuses on displaying the statistic and lacks additional space to represent the uncertainty. Providing an adjacent map or overlaying maps with symbols [@VSSDCUC] are two common solutions.

## 2.3 Limitations of choropleth displays {#ch:chorolimit}

@@

# 3 Contemporary alternatives to choropleth maps {#ch:alternatives}

## 3.1 Cartograms
<!-- 1496 words inc table-->

@@

<!-- Denominators -->
When a map creator desires a uniform population density of the map base, the purposeful distortion of the map space is beneficial. The "population distribution is often extremely uneven", making a distortion necessary so that population is more faithfully represented as a uniformly distributed background for the statistic to be presented [@ACTUC] [@CTTMB] [@GOINO]. An area cartogram [@NAC], or population-by-area cartogram [@TAAM] is produced from the distortion of the geographical shape according to population. Event cartograms [@VSSDCUC] change the area of regions on a map depending on the amount of disease-related events, rather than population.

<!-- Why transform?-->
<!-- Common variables used to create cartograms, e.g. population, mortality -->
Cartograms provide an alternative visualization method for statistical and geographical information. Monmonier [@HTLWM] suggests that map creators can use white lies to create useful spatial displays. It is easy for the reader to disregard the impact of transformations used to create cartograms, for the benefit of reading the statistical distribution more accurately with approximate geographic information. The spatial transformation of map regions relative to the data emphasizes the data distribution instead of land size [@CBATCC]. When visualizing population statistics, Dorling  considers this design 'more socially just' [@ACTUC], or honest [@NISCC], giving equitable representation and attention to all members of the population and reducing the visual impact of large areas with small populations [@DMAHP]. Howe [@HEDP] suggests that 'cancer occurs in people, not in geographical areas' and that spatial socio-economic data, like cancer rates, are best presented on a cartogram for urban areas as the population map base avoids allocating 'undue prominence' to rural areas [@CTTMB].

@@

```{r ggcartograms, fig.cap = "Figure 2: Common alternatives to maps, showing the same information for the United States of America: (a) contiguous cartogram, (b) non-contiguous, shape-preserved cartogram, (c)  Dorling cartogram (non-contiguous), (d) hexagon tilemap (non-contiguous). Maps (a) - (c) are created by resizing and reshaping the states of the USA to match the 2015 population of the state. This provides a better sense of the extent of disease relative to the population in the country and can help ease losing information about physically small but population-dense states. Map creators give each state equal size and thus equal emphasis in (d) the hexagon tile map.", message=FALSE, warning=FALSE, out.width='100%'}
g1 <- png::readPNG("figures/ggcont.png")
g2 <- png::readPNG("figures/ggncont.png")
g3 <- png::readPNG("figures/ggdorl.png")
g4 <- png::readPNG("figures/gghexmap.png")
#knitr::include_graphics("figures/usa_grid.png")

@@

Table: Table 3: <!--\label{tab:usa}--> Maps used to present statistics for the United States of America. The colour of each state communicates the average age-adjusted rate of incidence for lung and bronchus for females and males in the United States 2012-2016.

@@

### 3.1.1 Contiguous

@@

A contiguous cartogram alters the choropleth according to a statistic and maintains connectivity of the map regions. Min Ouyang and Revesz [@ACA] present three algorithms for creating value-by-area cartograms. They implement 'map deformation' to account for the value assigned to each area. Other methods include Tobler’s Pseudo-Cartogram Method, Dorling’s Cellular Automaton Method [@ACTUC], Radial Expansion Method, Rubber Sheet Method, Line Integral Method, Constraint-Based Method [@CBATCC].

Figure 2a<!--\ref{fig:ggcartograms}--> shows a population contiguous cartogram of the United States. All states are visible and the shape of the United States overall is still recognizable. In contrast, Figure 3<!--\ref{fig:auscartograms}--> a shows an Australian contiguous cartogram also based on population. The south east is enlarged, but high population areas are still small, and low population areas are still large on the map. The algorithm doesn't fully reach an optimal configuration where area matches population -- Australia is too heterogeneous for the algorithm to handle.

@@

### 3.1.2 Non-contiguous

Non-contiguous cartograms prioritize the shapes of the areas instead of connectivity. Each area stays in a similar position to its location on a choropleth map. Displaying the choropleth map base allows map users to make comparisons regarding the change in the area. The addition is the gap between areas, created as each area shrinks or grows according to the associated value of the statistic. Olson [@NAC] discusses the creation of these maps and the significance of the empty areas left between the geographic boundaries and the new shape.

The white space presents the meaningful empty-space property [@ECGC] [@NAC] but it also distracts the reader from the data, with a low data density [@tufte].

### 3.1.3 Dorling

Daniel Dorling presents an alternative display engineered to highlight the spatial distribution and neighborhood relationships without complex distortions of borders and boundaries [@ACTUC]:

@@

He acknowledged the sophistication of contiguous cartograms but critiqued their 'very complex shapes,' he answers this with his implementation of maps created using 'the simplest of all shapes'. Circular cartograms use the same circle shape for every region represented, resized  according to the statistic represented or the population. This simple shape may be more effective for understanding the spatial distribution than contiguous cartograms. Contiguous cartograms create 'nonsense' shapes that have 'no meaning' [@NISCC]. Both methods applies a gravity model to produce a layout, that avoids overlaps and keep spatial relationships with neighboring areas over many iterations. The circular cartogram is relatively fast to compute.

Raisz [@RSCW] laid the groundwork for this approach in the mid-1930s, drawing rectangular cartograms that provide simple comparisons, effective for correcting misconceptions communicated by geographic maps. Tobler @TFYCC names and defines these as Value-Area Cartograms. This rectangular display may sacrifice contiguity but allows for tiling where geographic neighbors placed in suitable relative positions also share borders [@CDWCS]. Rectangular cartograms communicate bivariate displays of the population by the size of each rectangular, and they use color to communicate a second variable [@ORC].

## 3.2 Tile Map
<!-- 173 words -->
A tile map provides a tessellated display of consistent shapes.  A similar method to a rectangular cartogram, represents each geographic area using a square. The squares are tessellated to create a grid. Each area is represented by a square of the same dimensions, each tile is usually one unit of measurement. This could be geographic regions such as states or population-based that use a consistent measure of population for each tile. Regions with more than four neighbors introduce some displacement, because only four squares can be adjoining. Like the chorpleth and cartograms, the tile map uses color to represent a value of a statistic for each area. There are online media sources using this method, which include [NPR](https://www.npr.org/2016/10/18/498406765/npr-battleground-map-hillary-clinton-is-winning-and-its-not-close), [FiveThirtyEight](https://projects.fivethirtyeight.com/2016-election-forecast/), [The Wall Street Journal](http://graphics.wsj.com/elections/2016/2016-electoral-college-map-predictions/), [The Washington Post](https://www.washingtonpost.com/graphics/politics/2016-election/50-state-poll/). Tile maps often need to be created manually, requiring more time and care as the number of geographic areas increases.
<!-- cite mosaic cartograms -->

Cano and others [@MDAC] define the term 'mosaic cartograms' for hexagonal tile displays, where the number of tiles for each area or the color of them can communicate the statistic of regions. When using several tiles per region, map makers can adjust the complexity of the boundaries in the resulting display. They can also make a trade-off between boundary complexity and simplicity by the size of the tiles used.
